$\pi$ with leftovers: \\ a mechanisation in Agda

### 2020-07-13 paper/paper.tex
First batch of mostly small changes addressing reviews

@@

% Functions
\newcommand{\subst}[3]{#1 \, \func{[} \, #3 \, \func{\mapsto} \,#2 \, \func{]}}
\newcommand{\op}[3]{#1 \, \func{\coloneqq} \, #2 \, \func{\cdot} \, #3}
\newcommand{\opsquared}[3]{#1 \, \func{\coloneqq} \, #2 \, \func{\cdot^2} \, #3}
\newcommand{\opctx}[3]{#1 \, \func{\coloneqq} \, #2 \, \func{\otimes} \, #3}
\newcommand{\zero}{\func{0\cdot}}
\newcommand{\one}{\func{1\cdot}}
\newcommand{\li}{\func{\ell_i}}
\newcommand{\lo}{\func{\ell_o}}
\newcommand{\lz}{\func{\ell_{\o}}}
\newcommand{\lio}{\func{\ell_{\#}}}

@@

We live in a concurrent world where any given state is often interactively computed by a myriad of parties --- people, machines, processors, services, networks.
As humans, we aim to model, predict, and build such interactive systems; as mathematicians, abstraction is our tool of choice to do so.

The \picalc{} \cite{MilnerPW92,Milner99} is the most successful computational model for communication and concurrency.
It abstracts over the details of concurrent processing and boils the interactions down to the sending and receiving of data over communication channels.
Notably, it features channel mobility: channels themselves are first order values, and can therefore be transmitted.
The \picalc{} has been a foundation for the design and implementation of programming languages for concurrency, such as Pict \cite{PierceT00} and (more recently) Go \footnote{https://golang.org}.
At the state-of-the-art, the \picalc{} features a wide plethora of types \cite{K07}: basic types, linear types, types for liveness properties (such as deadlock freedom, livelock freedom or termination), and most notably, session types. This makes the \picalc{} a fully-fledged tool for modelling and verification of concurrent and distributed systems.

@@

We do so by defining a specification for an algebra on \emph{multiplicities} (how many times a channel can be used) and use it to apply leftover typing (following Allais \cite{Allais2018a}) to the \picalc{} for the first time.
The user is able to choose any mix of algebras for the type system --- as long as they comply with the specification.
This allows for linear, graded and shared types to be used either separately or mixed in a type system.
Ultimately, we can exploit this formalisation as a unified framework for type safe distributed modelling and programming with a variety of type systems, and build on top of it more advanced types, theories and languages.
In particular, while linearity is needed for type safety in communication-centric programming, graded and shared types are needed to model real-world software systems.

\subsection{Contributions}
\begin{enumerate}
\item \textbf{Formalisation of the syntax and the semantics of the \picalc{}}:
 \begin{itemize}
   \item \textbf{Syntax}: \autoref{syntax} uses type level \emph{de Bruijn indices} \cite{deBruijn1972, Dybjer1994} to introduce a syntax that is \emph{well scoped by construction}: every free variable is accounted for in the type of the process that uses it.

@@

\autoref{fig:notation} illustrates the notation used in this paper.
Data type definitions ($\type{\N}$) use double lines and index-free synonyms (\textsc{Nat}) as rule names (for ease of reference).
Constructors ($\constr{0}$ and $\suc$) are given as inference rules, although we occasionally use a BNF style definition of types for brevity.
We maintain a close correspondence between the definitions presented in this paper and our mechanised definitions in Agda --- where premises become argument types and conclusions return types.
Universe levels and universe polymorphism are omitted for brevity --- all our types are of type $\Set$.
Implicit arguments are mentioned by type definitions but omitted by constructors.

We use colours to further distinguish the different entities in this paper.
$\type{TYPES}$ are blue violet (uppercased, with indices as subscripts), $\constr{constructors}$ are burnt orange, $\func{functions}$ are olive green, variables are black, and some constructor names are overloaded --- and disambiguated by context.

@@

\begin{proof}
  TODO
\end{proof}

@@

\begin{nidefinition}[Usage algebra]
  A \emph{usage algebra} is a ternary relation $\op{x}{y}{z}$, that is \emph{partial}, \emph{decidable}, \emph{deterministic}, \emph{cancellative}, \emph{associative}, and \emph{commutative}.
  It has a \emph{neutral element} $\zero$ that is absorbed on either side, and that is also \emph{minimal}.
  It has an element $\one$ that is used to count inputs and outputs.
  \autoref{fig:multiplicities} uses a ternary relation to model such an algebra as a record $\Algebra_C$ on a carrier $C$.

@@

Linear, graded and shared types are all defined as an instance of our usage algebra.
Their use in typing derivations is illustrated in \autoref{example-derivations}.
\begin{description}
  \item [Linear]
    The carrier is a type with two trivial constructors $\constr{zero}$ and $\constr{one}$.
    The monoidal operation has the element $\constr{zero}$ as neutral on both sides, and the element $\constr{one}$ splitting into $\constr{one}$ and $\constr{zero}$ (or $\constr{zero}$ and $\constr{one}$), and is uninhabited otherwise.
    All the properties of the algebra follow trivially.
    As an example, the multiplicity pair $\ell_i$ is represented as $\constr{one} \comma \constr{zero}$.

    \item [Graded]
    The carrier is the type of natural numbers.
    The element $\zero$ corresponds to 0, and $\one$ to 1.
    The partial monoid $\op{x}{y}{z}$ is defined exactly when $x = y + z$.
    All other properties of the algebra follow from the algebraic rules for the addition of natural numbers.
    As an example, we use $5 \comma 3$ to represent the multiplicities of a channel that is able to input 5 times and output 3 times.

    \item [Shared]
    The carrier is a type with a single trivial constructor $\constr{\omega}$.
    Both $\zero$ and $\one$ are instantiated to $\constr{\omega}$.
    The relation $\op{\constr{\omega}}{\constr{\omega}}{\constr{\omega}}$ is inhabited.
    All the properties of the algebra follow trivially.
    Shared channels have multiplicities $\constr{\omega} \comma \constr{\omega}$, which they will forever preserve.
\end{description}

@@

We use indexed sets of usage algebras to allow several usage algebras to coexist in our type system with leftovers (\autoref{leftover-typing}).
\begin{nidefinition}[Indexed set of usage algebras]
  An \emph{indexed set of usage algebras} is a nonempty type $\Idx$ of indices together with an interpretation $\Usage$ of indices into types, and an interpretation $\type{ALGEBRAS}$ of indices into usage algebras (\autoref{fig:indexed-multiplicities}).
  \begin{figure}[h]
    \begin{equation*}
      \begin{aligned}
        &\Idx               &:{} &\Set \\
        &\type{\exists IDX} &:{} &\Idx \\
        &\Usage             &:{} &\Idx \to \Set \\
        &\type{ALGEBRAS}    &:{} &(idx : \Idx) \to \Algebra_{\Usage_{idx}}
      \end{aligned}
    \end{equation*}
    \caption{Indexed set of usage algebras.}
    \label{fig:indexed-multiplicities}
  \end{figure}
\end{nidefinition}

@@

\begin{proof}[Proof (Sketch)]
  All the above theorems are proved by induction on \textsc{Process} and \textsc{Var}.
  For details, refer to our mechanisation in Agda \cite{Zalakain2020Agda}.
\end{proof}

@@

Allais \cite{Allais2018a} uses leftover typing to mechanise in Agda a bidirectional type system for the linear \lambdacalc{}.
He proves type preservation and provides a decision procedure for type checking and type inference.
In our paper, we apply leftover typing to the \picalc{} following Allais \cite{Allais2018a}, and focus on type checking. We give a more general usage algebra, leading to linear, graded and shared type systems.
Drawing from quantitative type theory (by McBride and Atkey \cite{McBride2016, Atkey2018}), in our work, we too are able to talk about fully consumed resources.
As such, we can transmit over a channel the $\lz$ multiplicities of a fully exhausted channel.
%
In their recent paper (unpublished at the time of writing), Ciccone and Padovani mechanise a dependently-typed linear \picalc{} in Agda \cite{Ciccone}.
They do so by defining an intrinsicly-typed calculus, which allows them to leverage Agda to provide a dependently-typed interpretation of messages.
Contexts contain not only types, but also the messages that inhabit their corresponding interpretation in Agda.
To avoid linearity violations, channel types as interpreted as unit types --- and therefore only present in the context in the object language.
This approach allows them to encode message predicates, branching, and variable-length conversations, as message input is modelled as dependent function abstraction in Agda.
They use the concrete set of multiplicities $0$, $1$, $\omega$, define a partial algebra on contexts and require extrinsic proofs of context split.
%
Orchard et al. introduce Granule \cite{Orchard}, a fully-fledged functional language with graded modal types, linear types, indexed types and polymorphism.
Modalities include exact usages, security levels and intervals; resource algebras are pre-ordered semirings with partial addition.
They provide bidirectional typing rules, and show the type safety of their semantics.

@@

Further work on mechanising the \picalc{} \cite{Henry-Gerard1999, Honsell2001a, Bengtson2013, Despeyroux2000, Affeldt2008}, focuses on  non-linear variations, whether we present a range of linear, graded and shared types.

>>>>

% Newline inside table
\usepackage{makecell}

% Fix autoref's example treatment 
\newtheorem{example}{Example}
\newcommand{\exampleautorefname}{Example}

@@

% Functions
\newcommand{\subst}[3]{#1 \, \func{[} \, #3 \, \func{\mapsto} \,#2 \, \func{]}}
\newcommand{\op}[3]{#1 \, \func{\coloneqq} \, #2 \, \func{\cdot} \, #3}
\newcommand{\opsquared}[3]{#1 \, \func{\coloneqq} \, #2 \, \func{\cdot^2} \, #3}
\newcommand{\opctx}[3]{#1 \, \func{\coloneqq} \, #2 \, \func{\otimes} \, #3}
\newcommand{\zero}{\func{0\cdot}}
\newcommand{\one}{\func{1\cdot}}
\newcommand{\li}{\func{\ell_i}}
\newcommand{\lo}{\func{\ell_o}}
\newcommand{\lz}{\func{\ell_{\varnothing}}}
\newcommand{\lio}{\func{\ell_{\#}}}

@@

We live in a concurrent world where any given state is often interactively computed by a myriad of parties --- people, machines, processors, services, networks.
As humans, we aim to model, build, and predict such interactive systems; as mathematicians, abstraction is our tool of choice to do so.

The \picalc{} \cite{MilnerPW92,Milner99} is the most successful computational model for communication and concurrency.
It abstracts over the details of concurrent processing and boils the interactions down to the sending and receiving of data over communication channels.
Notably, it features channel mobility: channels themselves are first class values, and can therefore be transmitted.
The \picalc{} has been a foundation for the design and implementation of programming languages for concurrency, such as Pict \cite{PierceT00} and (more recently) Go \footnote{https://golang.org}.
At the state-of-the-art, the \picalc{} features a wide plethora of types \cite{K07}: basic types, linear types, types for liveness properties (such as deadlock freedom, livelock freedom or termination), and most notably, session types. This makes the \picalc{} a fully-fledged tool for modelling and verification of concurrent and distributed systems.

@@

We do so by defining a specification for an algebra on \emph{multiplicities} (how many times a channel can be used) and use it to apply leftover typing (following Allais \cite{Allais2018a}) to the \picalc{} for the first time.
The user is able to choose any mix of algebras for the type system --- as long as they comply with the specification.
This allows for linear, graded and shared types to be used either separately or mixed in a type system.
Ultimately, we can exploit this formalisation as a unified framework for type safe distributed modelling and programming with a variety of type systems, and build on top of it more advanced types, theories and languages.

\subsection{Contributions}
\begin{enumerate}
\item \textbf{Formalisation of the syntax and the semantics of the \picalc{}}:
 \begin{itemize}
   \item \textbf{Syntax}: \autoref{syntax} uses type level \emph{de Bruijn indices} \cite{deBruijn1972, Dybjer1994} to introduce a syntax that is \emph{well scoped by construction}: every free variable is accounted for in the type of the process that uses it.
     For convenience, the user can convert from and to a syntax with names.
     These two conversion functions are shown to be inverses of one another modulo $\alpha$-conversion.

@@

\autoref{fig:notation} illustrates the notation used in this paper.
Data type definitions ($\type{\N}$) use double lines and index-free synonyms (\textsc{Nat}) as rule names (for ease of reference).
Constructors ($\constr{0}$ and $\suc$) are given as inference rules, although we occasionally use a BNF style definition of types for brevity.
We maintain a close correspondence between the definitions presented in this paper and our mechanised definitions in Agda --- where premises become argument types and conclusions return types.
Universe levels and universe polymorphism are omitted for brevity --- all our types are of type $\Set$.
Implicit arguments are mentioned in type definitions but omitted by constructors.

We use colours to further distinguish the different entities in this paper.
$\type{TYPES}$ are blue (uppercased, with indices as subscripts), $\constr{constructors}$ are orange, $\func{functions}$ are olive green, variables are black, and some constructor names are overloaded --- and disambiguated by context.

@@

\begin{proof}[Sketch]
  All the above theorems are proved by induction on \textsc{Process} and \textsc{Var}.
  For details, refer to our mechanisation in Agda \cite{Zalakain2020Agda}.
\end{proof}

@@

\begin{nidefinition}[Usage algebra]
  A \emph{usage algebra} is a ternary relation $\op{x}{y}{z}$, that is \emph{partial}, \emph{decidable}, \emph{deterministic}, \emph{cancellative}, \emph{associative}, and \emph{commutative}.
  It has a \emph{neutral element} $\zero$ that is absorbed on either side, and that is also \emph{minimal}.
  It has an element $\one$ that is used to count inputs and outputs.
  \autoref{fig:multiplicities} models such an algebra as a record $\Algebra_C$ on a carrier $C$.

@@

\autoref{fig:algebra-instances} sketches the implementation of linear, graded and shared types as instances of our usage algebra.
Their use in typing derivations is illustrated in \autoref{example-derivations}.

\begin{figure}[h]
\begin{tabular}{l | l | l}
  & carrier & operation \\
  \hline
  \textbf{linear} & \makecell[cl]{$\constr{0} \, : \, \type{Lin}$ \\ $\constr{1} \, : \, \type{Lin}$} & \makecell[cl]{$\op{\constr{0}}{\constr{0}}{\constr{0}}$ \\ $\op{\constr{1}}{\constr{1}}{\constr{0}}$ \\ $\op{\constr{1}}{\constr{0}}{\constr{1}}$} \\
  \hline
  \textbf{graded} & \makecell[cl]{$\constr{0} \, : \type{Gra}$ \\ $\suc \, : \type{Gra} \to \type{Gra}$} & \makecell[cl]{$\forall \, x \, y \, z$ \\ $\to x \, \type{\equiv} \, y \, \func{+} \, z$ \\ $\to \op{x}{y}{z}$} \\
  \hline
  \textbf{shared} & $\constr{\omega} \, : \, \type{Sha}$ & $\op{\constr{\omega}}{\constr{\omega}}{\constr{\omega}}$ \\
\end{tabular}
\caption{Example instances of $\Usage$ and a ternary relation that satisfies our $\Algebra$ interface.}
\label{fig:algebra-instances}
\end{figure}

@@

We use indexed sets of usage algebras to allow several usage algebras to coexist in our type system with leftovers (\autoref{leftover-typing}).
\begin{nidefinition}[Indexed set of usage algebras]
  An \emph{indexed set of usage algebras} is a type $\Idx$ of indices that is nonempty ($\type{\exists IDX}$) together with an interpretation $\Usage$ of indices into types, and an interpretation $\type{ALGEBRAS}$ of indices into usage algebras of the corresponding type (\autoref{fig:indexed-multiplicities}).
  \begin{figure}[h]
    \begin{equation*}
      \begin{aligned}
        &\Idx               &:{} &\Set \\
        &\type{\exists IDX} &:{} &\Idx \\
        &\Usage             &:{} &\Idx \to \Set \\
        &\type{ALGEBRAS}    &:{} &(idx : \Idx) \to \Algebra_{\Usage_{idx}}
      \end{aligned}
    \end{equation*}
    \caption{Indexed set of usage algebras.}
    \label{fig:indexed-multiplicities}
  \end{figure}
\end{nidefinition}

@@

\begin{proof}[Proof (Sketch)]
  All the above theorems are proved by induction on \textsc{Types} and \textsc{VarRef}.
  For details, refer to our mechanisation in Agda \cite{Zalakain2020Agda}.
\end{proof}

@@

Allais \cite{Allais2018a} uses leftover typing to mechanise in Agda a bidirectional type system for the linear \lambdacalc{}.
He proves type preservation and provides a decision procedure for type checking and type inference.
In this paper, we follow Allais \cite{Allais2018a} and apply leftover typing to the \picalc{}.
We give a more general usage algebra, leading to linear, graded and shared type systems.
Drawing from quantitative type theory (by McBride and Atkey \cite{McBride2016, Atkey2018}), in our work, we too are able to talk about fully consumed resources --- e.g., we can transmit over a channel $\lz$ multiplicities of a fully exhausted channel.
%
In their recent paper (unpublished at the time of writing), Ciccone and Padovani mechanise a dependently-typed linear \picalc{} in Agda \cite{Ciccone}.
They do so by defining an intrinsicly-typed calculus, which allows them to leverage Agda to provide a dependently-typed interpretation of messages.
Contexts contain not only types, but also the messages that inhabit their corresponding interpretation in Agda.
To avoid linearity violations, channel types as interpreted as unit types --- and therefore only present in the context in the object language.
This approach allows them to encode message predicates, branching, and variable-length conversations, as message input is modelled as dependent function abstraction in Agda.
They use the concrete set of multiplicities $0$, $1$, $\omega$, define a partial algebra on contexts and require extrinsic proofs of context split.
%
Orchard et al. introduce Granule \cite{Orchard}, a fully-fledged functional language with graded modal types, linear types, indexed types and polymorphism.
Modalities include exact usages, security levels and intervals; resource algebras are pre-ordered semirings with partial addition.
They provide bidirectional typing rules, and show the type safety of their semantics.

@@

Further work on mechanising the \picalc{} \cite{Henry-Gerard1999, Honsell2001a, Bengtson2013, Despeyroux2000, Affeldt2008}, focuses on non-linear variations, whereas we present a range of linear, graded and shared types.
